---
title: "Assignment 2 - Language Development in ASD - Part 1 - Explaining development"
author: "Mie Arnau Martinez"
date: "12.09.2019"
output:   
  md_document:
    variant: markdown_github
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = FALSE)
```

# Assignment 2

In this assignment you will have to discuss a few important questions (given the data you have). More details below. The assignment submitted to the teachers consists of:
- a report answering and discussing the questions (so we can assess your conceptual understanding and ability to explain and critically reflect)
- a link to a git repository with all the code (so we can assess your code)

Part 1 - Basic description of language development
- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
- Include individual differences in your model of language development (in children). Identify the best model.

Part 2 - Model comparison
- Discuss the differences in performance of your model in training and testing data
- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
- Predict a new kid's performance (Bernie) and discuss it against expected performance of the two groups

Part 3 - Simulations to plan a new study
- Report and discuss a power analyses identifying how many new kids you would need to replicate the results

The following involves only Part 1.

## Learning objectives

- Summarize and report data and models
- Critically apply mixed effects (or multilevel) models
- Explore the issues involved in feature selection


# Quick recap
Autism Spectrum Disorder is often related to language impairment. However, this phenomenon has not been empirically traced in detail:
i) relying on actual naturalistic language production,  ii) over extended periods of time.

We therefore videotaped circa 30 kids with ASD and circa 30 comparison kids (matched by linguistic performance at visit 1) for ca. 30 minutes of naturalistic interactions with a parent. We repeated the data collection 6 times per kid, with 4 months between each visit. We transcribed the data and counted: 
i) the amount of words that each kid uses in each video. Same for the parent.
ii) the amount of unique words that each kid uses in each video. Same for the parent.
iii) the amount of morphemes per utterance (Mean Length of Utterance) displayed by each child in each video. Same for the parent. 

This data is in the file you prepared in the previous class. 

NB. A few children have been excluded from your datasets. We will be using them next week to evaluate how good your models are in assessing the linguistic development in new participants.

This RMarkdown file includes 
1) questions (see above). Questions have to be answered/discussed in a separate document that you have to directly send to the teachers.
2) A break down of the questions into a guided template full of hints for writing the code to solve the exercises. Fill in the code and the paragraphs as required. Then report your results in the doc for the teachers.

REMEMBER that you will have to have a github repository for the code and send the answers to Kenneth and Riccardo without code (but a link to your github/gitlab repository). This way we can check your code, but you are also forced to figure out how to report your analyses :-)

Before we get going, here is a reminder of the issues you will have to discuss in your report:

1- Describe your sample (n, age, gender, clinical and cognitive features of the two groups) and critically assess whether the groups (ASD and TD) are balanced
2- Describe linguistic development (in terms of MLU over time) in TD and ASD children (as a function of group). 
3- Describe how parental use of language (in terms of MLU) changes over time. What do you think is going on?
4- Include individual differences in your model of language development (in children). Identify the best model.

# Let's go

### Loading the relevant libraries

Load necessary libraries : what will you need?
- e.g. something to deal with the data
- e.g. mixed effects models
- e.g. something to plot with

```{r Load Libraries, include = FALSE}
library(pacman)
p_load(tidyverse, lme4, lmerTest, MuMIn)
```

### Define your working directory and load the data
If you created a project for this class and opened this Rmd file from within that project, your working directory is your project directory.

If you opened this Rmd file outside of a project, you will need some code to find the data:
- Create a new variable called locpath (localpath)
- Set it to be equal to your working directory
- Move to that directory (setwd(locpath))
- Load the data you saved last time (use read_csv(fileName))

```{r Load Data, include = FALSE}
# Loading data
train <- read.csv("train.csv", header = T, sep = ",")
```

### Characterize the participants (Exercise 1)

Identify relevant variables: participants demographic characteristics, diagnosis, ADOS, Verbal IQ, Non Verbal IQ, Socialization, Visit, Number of words used, Number of unique words used, mean length of utterance in both child and parents.

Make sure the variables are in the right format.

Describe the characteristics of the two groups of participants and whether the two groups are well matched.

```{r descriptive stats, include = FALSE}
train$SUBJ <- as.character(train$SUBJ)
summary(train)

# Checking gender balance in each diagnosis as well as mean age counted in months
train %>%
  subset(VISIT == 1) %>%
  group_by(Diagnosis) %>%
  dplyr::summarise("number" = n(), 
                   "females" = sum(Gender == "F"), 
                   "males" = sum(Gender == "M"),
                   "AgeM" = mean(Age, na.rm = T),
                   "verbal-IQ" = mean(ExpressiveLangRaw1),
                   "non-verbal-IQ" = mean(MullenRaw1),
                   "ADOS" = mean(ADOS1),
                   "Caucasian" = sum(Ethnicity == "White"),
                   "Non Caucasian" = n() - sum(Ethnicity == "White"),
                   "socialization" = mean(Socialization1),
                   "number-words" = mean(tokens_CHI),
                   "unique-words" = mean(types_CHI),
                   "mlu_chi" = mean(CHI_MLU),
                   "mlu_mot" = mean(MOT_MLU))

```

The sample included mostly young (<20) white males ...

[REPORT THE RESULTS]

## Let's test hypothesis 1: Children with ASD display a language impairment  (Exercise 2)

### Hypothesis: The child's MLU changes: i) over time, ii) according to diagnosis

Let's start with a simple mixed effects linear model

Remember to plot the data first and then to run a statistical test.
- Which variable(s) should be included as fixed factors?
- Which variable(s) should be included as random factors?

```{r ex2, include = FALSE}
# Plot showing average intercept and language development for each diagnosis
plot1 <- ggplot(train, aes(VISIT, CHI_MLU, group = Diagnosis, color = Diagnosis))+
  geom_smooth(method = lm)+
  geom_point()+
  labs(title = "Development of mean length of utterance for ASD and TD over visits") +
  xlab("Visit") + ylab("Mean length utterance")

# Plot showing individual language development for each diagnosis
plot2 <- ggplot(train, aes(VISIT, CHI_MLU, group = SUBJ, color = Diagnosis))+
  geom_smooth(method = lm, se = FALSE)+
  facet_wrap(~Diagnosis)+
  theme_minimal() +
  labs(title = "Development of mean length of utterance for ASD and TD over visits") +
  xlab("Visit") + ylab("Mean length utterance")

# Slope of visit will depend on diagnosis. Differ between ASD and TD. (1|SUBJ) = every child is unique and their clusters of data are therefore different. Uniqueness is only related to the intercept. (0+VISIT|SUBJ) = Slope is different for each child.

# Mixed effect model
model1 <- lmerTest::lmer(data = train, CHI_MLU ~ VISIT*Diagnosis + (1|SUBJ) + (0+VISIT|SUBJ)) 

# Null model without the interaction effect
nullModel <- lmerTest::lmer(data = train, CHI_MLU ~ VISIT + Diagnosis + (1|SUBJ) + (0+VISIT|SUBJ))

```

How would you evaluate whether the model is a good model?

```{r ex2 evaluate, include = FALSE}
# Model coefficients for model 1 and null model
summary(model1)
summary(nullModel)

# Checking normality of residuals
qqnorm(residuals(model1))
# Since they the residuals of the model are normally distributed the model complies to the assumption of normality of residuals.

# Testing the model with analysis of variance
anova(model1, nullModel)

# Calculating marginal and conditional R-squared.
r.squaredGLMM(model1)

# From this we gather that there was a significant interaction effect of Diagnosis and visit on the mean length of utterance. However as can be seen from the calculation of R^2, variance of individual slopes and intercepts explains a lot more than just the interaction of diagnosis and visit. If we wanted to generalize the results to the general population this would cause problems.
```

Not too good, right? Let's check whether a growth curve model is better.
Remember: a growth curve model assesses whether changes in time can be described by linear, or quadratic, or cubic (or... etc.) components.
First build the different models, then compare them to see which one is better.

```{r ex2 growth curve, include = FALSE}

```

Exciting right? Let's check whether the model is doing an alright job at fitting the data. Plot the actual CHI_MLU data against the predictions of the model fitted(model). 

```{r}
# Fitting model to the data
df <- train
df$fit <- predict(model1) # Making a new column with the predicted mlu based on the model

# Predicted versus actual, points are actual datapoints, lines are predictions
ggplot(df, aes(x = VISIT, y = CHI_MLU, group = SUBJ, color = Diagnosis)) +
  geom_point(alpha = 0.5, position = position_jitter(w = 0.1, h = 0)) +
  facet_wrap(~SUBJ) +
  theme_minimal() +
  geom_smooth(method = "lm", se = FALSE, size = 0.4, aes(y=fit)) +
  labs(title = "Model prediction on current data")
```

Now it's time to report our results.
Remember to report:
- the estimates for each predictor (beta estimate, standard error, p-value)
- A plain word description of the results
- A plot of your model's predictions (and some comments on whether the predictions are sensible)

[REPORT THE RESULTS]
Linguistic development of children MLU is affected by ... [COMPLETE]

## Let's test hypothesis 2: Parents speak equally to children with ASD and TD  (Exercise 3)

### Hypothesis: Parental MLU changes: i) over time, ii) according to diagnosis

```{r ex3, include = FALSE}
# Model 2 predicting parental mean length utterance assuming no interaction
model2 <- lmer(data = train, MOT_MLU ~ VISIT + Diagnosis + (1|SUBJ) + (0 + VISIT|SUBJ))
# Assuming that the intercept does not affect the slope

# Model 3 predicting parental mean length utterance assuming interaction of visit and diagnosis
model3 <- lmerTest::lmer(data = train, MOT_MLU ~ VISIT * Diagnosis + (1|SUBJ) + (0 + VISIT|SUBJ))

# Summary of models
summary(model2)
summary(model3)

# Comparison of models using anova
anova(model2,model3) # model 3 doesn't explain more variance compared to model 2. 

# Calculating the marginal and conditional R-squared
r.squaredGLMM(model2)
```

The marginal R2 for adults shows that visit and diagnosis explains about 20 % of the variance where the conditional R2 includes the random effects and explains about 65%. This suggests that time and diagnosis can affect the Parent MLU but this might not be the only thing.


### Adding new variables (Exercise 4)

Your task now is to figure out how to best describe the children linguistic trajectory. The dataset contains a bunch of additional demographic, cognitive and clinical variables (e.g.verbal and non-verbal IQ). Try them out and identify the statistical models that best describes your data (that is, the children's MLU). Describe how you selected the best model and send the code to run the model to Riccardo and Kenneth


```{r ex4, include = FALSE}

```

In addition to ..., the MLU of the children is also correlated with ...
Using AIC / nested F-tests as a criterium, we compared models of increasing complexity and found that ...

[REPORT THE RESULTS]